# Yearly value

There is great variation in the value of the commodities produced over the years and among the minerals/metals. These dynamics are the focus of this chapter.

Value is $price \times quantity$.

```{r}
#| label: setup
#| include: false
#| warning: false

source("./scripts/setup-prepare-data.R")
# creates dta_yearly_long
# creates cpi

year_first <- min(dta_yearly_long$year)
year_last <- max(dta_yearly_long$year)
year_n = year_last - year_first + 1

peak_production_cutoff <- 0.9 # at least this much of amount_max to be considered peak
peak_price_cutoff <- 0.9 # at least this much of amount_max to be considered peak

```

<br>

## Contributions in 2021 to total value

Using log scale since the range is very wide.

```{r, fig.height=7, fig.width=12}
#| label: fig-value-2021
#| fig-cap: "Value extracted in 2021"
#| fig-width: 12
#| fig-height: 7
#| column: page-inset-right

product_levels <- dta_yearly_long |>
  filter(year == 2021,
         !is.na(value_2021)) |>
  slice_max(order_by = value_2021, n = 1,
            by = product_name) |>
  arrange(value_2021)

p1 <- dta_yearly_long |>
  filter(year == 2021) |>
  inner_join(product_levels |> select(product_name),
             by = "product_name") |>
  mutate(product_name = factor(product_name, levels = product_levels$product_name)) |>
  ggplot(aes(year, value_2021, label = product_name, color = product_name)) +
  geom_label_repel(na.rm = TRUE, show.legend = FALSE,
                   direction = "x", min.segment.length = 100) +
  scale_x_continuous(breaks = c(1900, 1950, 2000)) +
  scale_y_log10(labels = label_number(scale_cut = cut_short_scale(),
                                           prefix = "$"),
  ) +
  labs(
    subtitle = glue("Relative commodity value"),
    x = NULL,
    y = "AUD (log10 scale)"
  )

data_for_plot <- dta_yearly_long |>
  filter(year == 2021) |>
  inner_join(product_levels |> select(product_name),
             by = "product_name") |>
  arrange(desc(value_2021)) |>
  mutate(product_name = factor(product_name, levels = product_levels$product_name),
         cum_value = cumsum(value_2021))

max_cum_value = max(data_for_plot$cum_value)
  
p2 <- data_for_plot |>
  ggplot(aes(cum_value, product_name)) +
  #geom_path(aes(group = cum_value), alpha = 0.4, color = carolina_blue) +
  geom_point() +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale()),
                     sec.axis = sec_axis(~ . / max_cum_value,
                                         labels = label_percent(accuracy = 1))) +
  expand_limits(x = 0) +
  labs(
    subtitle = glue("Commodity contribution to total value in 2021 (AUD {round(max_cum_value / 1e9, digits = 0)}B)"),
    x = "Cumulative value (AUD)",
    y = NULL
  )

p1 + p2 +
  plot_annotation(
    title = 'Australian mining: value of products produced in 2021',
    caption = my_caption
  )

```

<br>

## Yearly value by commodity

```{r, fig.height=10, fig.width=14}
#| label: fig-value-over-time-2021-dollars
#| fig-cap: "Value over time in 2021 dollars"
#| fig-width: 14
#| fig-height: 10
#| column: screen-inset-right

product_levels <- dta_yearly_long |>
  filter(!is.na(value_2021)) |>
  slice_max(order_by = value_2021, n = 1,
            by = product_name) |>
  arrange(desc(value_2021)) |>
  pull(product_name)

dta_yearly_long |>
  mutate(product_name = factor(product_name, levels = product_levels)) |>
  ggplot(aes(year, value_2021, color = product_name)) +
  geom_line(na.rm = TRUE, show.legend = FALSE) +
  scale_x_continuous(breaks = c(1900, 1950, 2000)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(),
                                           prefix = "$"),
  ) +
  facet_wrap(~ product_name, scales = "free_y") +
  labs(
    title = glue("Australian mining yearly value of products (2021 dollars)"),
    subtitle = glue("Arranged from highest to lowest peak value. {year_first}-{year_last}"),
    x = NULL,
    y = "AUD (2021 dollars) Y axis scale varies",
    caption = my_caption
  )

```

<br>

## Commodities at or near peak value of production

For the following, the value of production in 2021 was within `r percent(1 - peak_production_cutoff)` of historical peak value.


```{r}
#| label: fig-output-yearly-at-peak
#| fig-cap: "Products at or near peak production: yearly Australian mining output"

peak_products <- dta_yearly_long |>
  filter(year == max(year, na.rm = TRUE) & value_2021_pct_of_max >= peak_production_cutoff) 

data_for_plot <- dta_yearly_long |>
  inner_join(peak_products %>% select(product_price),
             by = "product_price")

labels_for_plot <- data_for_plot |>
  filter(!is.na(mass)) |>
  mutate(final_year = max(year),
         .by = product_price) |>
  filter(year == max(year),
         .by = product_price)
  
data_for_plot |>
  ggplot(aes(year, mass, color = product_price, group = product_price)) +
  geom_line(na.rm = TRUE, show.legend = FALSE) +
  geom_text_repel(data = labels_for_plot,
                  aes(max(year) + 30, mass, label = product_price),
                  hjust = 1, vjust = 0.5, show.legend = FALSE, 
                  direction = "y", force = 0.4) +
  scale_x_continuous(breaks = c(1850, 1900, 1950, 2000)) +
  scale_y_log10(labels = label_number(scale_cut = cut_short_scale(),
                                      prefix = "$"),
  ) +
  labs(
    title = glue("Production near peak value in Australia"),
    subtitle = glue("{year_first}-{year_last}"),
    x = NULL,
    y = "AUD (2021 dollars) log10 scale",
    caption = my_caption
  )
  
```

<br>

## Table

```{r}
#| label: tbl-summary-value-2021
#| tbl-cap: "Australian mining value extracted in 2021"

dta_yearly_long |>
  filter(year == 2021) |> 
  select(product_name, units_price, value_2021, value_max_2021) |>
  mutate(value_pct_of_max = value_2021 / value_max_2021) |>
  arrange(desc(value_2021), desc(value_pct_of_max)) |>
  mutate(cum_pct = cumsum(value_2021 / sum(value_2021, na.rm = TRUE)),
         rowid = row_number()) |>
  gt() |>
  tab_header(md("**Australian mining value extracted 2021**<br>*Australian dollars (AUD)*")) |>
  fmt_currency(columns = c(value_2021, value_max_2021),
             decimals = 0,
             currency = "AUD",
             suffixing = TRUE) |>
  fmt_percent(columns = c(value_pct_of_max, cum_pct),
             decimals = 0) |>
  sub_missing() |>
  tab_source_note(md("*Data: Gavin Mudd. Analysis: Daniel Moul*"))

```

<br>
