# Yearly prices

There is great variation in commodity prices over time and among the metals/minerals.

<br>

```{r}
#| label: setup
#| include: false
#| warning: false

source("./scripts/setup-prepare-data.R")
# creates dta_yearly_long
# creates cpi

year_first <- min(dta_yearly_long$year)
year_last <- max(dta_yearly_long$year)
year_n = year_last - year_first + 1

peak_production_cutoff <- 0.9 # at least this much of amount_max to be considered peak
peak_price_cutoff <- 0.9 # at least this much of amount_max to be considered peak

```

A lot of metals/minerals experienced very high nominal price rises since 2000.

```{r, fig.height=10, fig.width=14}
#| label: fig-prices-over-time
#| fig-cap: "Prices over time"
#| fig-width: 14
#| fig-height: 10
#| column: screen-inset-right

product_levels <-dta_yearly_long |>
  filter(!is.na(price)) |>
  slice_max(order_by = price, n = 1,
            by = product_price) |>
  arrange(desc(price)) |>
  pull(product_price)

dta_yearly_long |>
  mutate(product_price = factor(product_price, levels = product_levels)) |>
  ggplot(aes(year, price, color = product_price)) +
  geom_line(na.rm = TRUE, show.legend = FALSE) +
  scale_x_continuous(breaks = c(1850, 1900, 1950, 2000)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(),
                                           prefix = "$"),
  ) +
  facet_wrap(~ product_price, scales = "free_y") +
  labs(
    title = glue("Australian mining prices"),
    subtitle = "Highest to lowest prices",
    x = NULL,
    y = "AUD (nominal) Y axis scale varies",
    caption = my_caption
  )

```

<br>

But after adjusting for inflation (using constant 2021 dollars) we can see that commodity prices generally are not near their peak, and some prices have been in long-term decline (Cobalt, Tungsten, Lead, Zinc, Alumina, Phosphate rock, and Bauxite).

Thus all plots starting with this one use constant 2021 dollars.

```{r, fig.height=10, fig.width=14}
#| label: fig-prices-over-time-2021-dollars
#| fig-cap: "Prices over time in 2021 dollars"
#| fig-width: 14
#| fig-height: 10
#| column: screen-inset-right

product_levels <- dta_yearly_long |>
  filter(!is.na(price_2021)) |>
  slice_max(order_by = price_2021, n = 1,
            by = product_price) |>
  arrange(desc(price_2021)) |>
  pull(product_price)

dta_yearly_long |>
  mutate(product_price = factor(product_price, levels = product_levels)) |>
  
  ggplot(aes(year, price_2021, color = product_price)) +
  geom_line(na.rm = TRUE, show.legend = FALSE) +
  scale_x_continuous(breaks = c(1850, 1900, 1950, 2000)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(),
                                           prefix = "$"),
  ) +
  facet_wrap(~ product_price, scales = "free_y") +
  labs(
    title = glue("Australian mining prices (2021 dollars)"),
    subtitle = "Highest to lowest volumne",
    x = NULL,
    y = "AUD (2021 dollars) Y axis scale varies",
    caption = my_caption
  )

```

<br>

Price by volume: does volume go up when prices go up (mine owners are incented to mine more)? Or do prices go down when volume goes up (there is a surplus leading to lower prices)? The answer is mixed. Yes, for some (Gold, Lithium, Chromium, Iron ore, Coal) but not others (Uranium, Vanadium, Bauxite).

```{r, fig.height=10, fig.width=14}
#| label: fig-price-by-volume-2021-dollars
#| fig-cap: "Price by volume in 2021 dollars"
#| fig-width: 14
#| fig-height: 10
#| column: screen-inset-right
#| warning: false

product_levels <- dta_yearly_long |>
  filter(!is.na(price_2021)) |>
  slice_max(order_by = price_2021, n = 1,
            by = product_name) |>
  arrange(desc(price_2021)) |>
  pull(product_name)

dta_yearly_long |>
  filter(!is.na(price)) |>
  mutate(product_name = factor(product_name, levels = product_levels)) |>
  ggplot(aes(volume, price_2021, color = year)) +
  geom_point(na.rm = TRUE, size = 0.5, show.legend = TRUE) +
  geom_smooth(se = FALSE, method = 'loess', formula = 'y ~ x', 
              na.rm = TRUE, linewidth = 0.5, color = "firebrick") + 
  scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(labels = label_number(scale_cut = cut_short_scale(),
                                           prefix = "$")) +
  facet_wrap(~ product_name, scales = "free") +
  theme(legend.position = "top") +
  labs(
    title = glue("Australian mining prices by volume (2021 dollars)"),
    subtitle = "Highest to lowest prices",
    x = "Yearly volume X axis scale varies (log10 scale)",
    y = "Price AUD (2021 dollars) Y axis scale varies (log10 scale)",
    caption = my_caption
  )

```

<br>
