# Yearly prices

There is great variation in commodity prices over time and among the commodities. These dynamics are the focus of this chapter. 

<br>

```{r}
#| label: setup
#| include: false
#| warning: false

source("./scripts/setup-prepare-data.R")
# creates dta_yearly_long
# creates cpi

year_first <- min(dta_yearly_long$year)
year_last <- max(dta_yearly_long$year)
year_n = year_last - year_first + 1

peak_production_cutoff <- 0.9 # at least this much of amount_max to be considered peak
peak_price_cutoff <- 0.9 # at least this much of amount_max to be considered peak

```

# Prices by commodity: nominal and constant 2021 dollars

Since 2000, very high nominal price rises occurred for many commodities.

```{r, fig.height=10, fig.width=14}
#| label: fig-prices-over-time
#| fig-cap: "Prices over time"
#| fig-width: 14
#| fig-height: 10
#| column: screen-inset-right

product_levels <-dta_yearly_long |>
  filter(!is.na(price)) |>
  slice_max(order_by = price, n = 1,
            by = product_price) |>
  arrange(desc(price)) |>
  pull(product_price)

dta_yearly_long |>
  mutate(product_price = factor(product_price, levels = product_levels)) |>
  ggplot(aes(year, price, color = product_price)) +
  geom_line(na.rm = TRUE, show.legend = FALSE) +
  scale_x_continuous(breaks = c(1850, 1900, 1950, 2000)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(),
                                           prefix = "$"),
  ) +
  facet_wrap(~ product_price, scales = "free_y") +
  labs(
    title = glue("Australian mining prices"),
    subtitle = "Highest to lowest prices",
    x = NULL,
    y = "AUD (nominal) Y axis scale varies",
    caption = my_caption
  )

```

<br>

But after adjusting for inflation (using constant 2021 dollars) we can see that commodity prices generally are not near their peak, and some prices have been in long-term decline (Tungsten, Alumina, and Bauxite).

::: {.callout-note}
For this reason, all plots starting with this one use constant 2021 dollars.
:::

```{r, fig.height=10, fig.width=14}
#| label: fig-prices-over-time-2021-dollars
#| fig-cap: "Prices over time in 2021 dollars"
#| fig-width: 14
#| fig-height: 10
#| column: screen-inset-right

product_levels <- dta_yearly_long |>
  filter(!is.na(price_2021)) |>
  slice_max(order_by = price_2021, n = 1,
            by = product_price) |>
  arrange(desc(price_2021)) |>
  pull(product_price)

dta_yearly_long |>
  mutate(product_price = factor(product_price, levels = product_levels)) |>
  
  ggplot(aes(year, price_2021, color = product_price)) +
  geom_line(na.rm = TRUE, show.legend = FALSE) +
  scale_x_continuous(breaks = c(1850, 1900, 1950, 2000)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(),
                                           prefix = "$"),
  ) +
  facet_wrap(~ product_price, scales = "free_y") +
  labs(
    title = glue("Australian mining prices (2021 dollars)"),
    subtitle = "Highest to lowest volumne",
    x = NULL,
    y = "AUD (2021 dollars) Y axis scale varies",
    caption = my_caption
  )

```

<br>

## Price by production volume

Does production volume go up when prices go up (mine owners are more motivated to produce more, or it becomes economical to mine previously marginal deposits)? Or do prices go down when volume goes up (perhaps demand is satisfied, leading to lower prices)? The answer appears to be mixed. For some (Gold, Lithium, Chromium, Iron ore) demand and production volume have been positively correlated in recent years. Some are negatively correlated (Uranium, Bauxite).

```{r, fig.height=10, fig.width=14}
#| label: fig-price-by-volume-2021-dollars
#| fig-cap: "Price by volume in 2021 dollars"
#| fig-width: 14
#| fig-height: 10
#| column: screen-inset-right
#| warning: false

product_levels <- dta_yearly_long |>
  filter(!is.na(price_2021)) |>
  slice_max(order_by = price_2021, n = 1,
            by = product_name) |>
  arrange(desc(price_2021)) |>
  pull(product_name)

dta_yearly_long |>
  filter(!is.na(price)) |>
  mutate(product_name = factor(product_name, levels = product_levels)) |>
  ggplot(aes(mass, price_2021, color = year)) +
  geom_point(na.rm = TRUE, size = 0.5, show.legend = TRUE) +
  geom_smooth(se = FALSE, method = 'loess', formula = 'y ~ x', 
              na.rm = TRUE, linewidth = 0.5, color = "firebrick") + 
  scale_x_log10(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_log10(labels = label_number(scale_cut = cut_short_scale(),
                                           prefix = "$")) +
  facet_wrap(~ product_name, scales = "free") +
  theme(legend.position = "top") +
  labs(
    title = glue("Australian mining prices by production volume (2021 dollars)"),
    subtitle = "Highest to lowest prices",
    x = "Yearly production volume X axis scale varies (log10 scale)",
    y = "Price AUD (2021 dollars) Y axis scale varies (log10 scale)",
    caption = my_caption
  )

```

<br>
